A comparison of design-based and model-based approaches for finite population spatial sampling and inference.

### 2022-04-18 inst/manuscript/manuscript.tex
minor review updates

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulated and real
    data from the United States Environmental Protection Agency's 2012
    National Lakes Assessment. A variety of sample sizes, location
    layouts, dependence structures, and response types are considered. The
    population mean is the parameter of interest and performance is
    measured using statistics like bias, squared error, and interval
    coverage.
  \item
    When studying the simulated and real data, we found that regardless of
    the strength of spatial dependence in the data, the generalized random
    tessellation stratified (GRTS) algorithm, which explicitly
    incorporates spatial locations into sampling, tends to outperform the
    simple random sampling (SRS) algorithm, which does not explicitly
    incorporate spatial locations into sampling. We also found that
    model-based inference tends to outperform design-based inference, even
    for skewed data where the model-based distributional assumptions are
    violated. The performance gap between these design-based inference and
    model-based inference is small GRTS samples are used but large when
    SRS samples are used. This suggests that the sampling choice (whether
    to use GRTS or SRS) is most important when performing design-based
    inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial sampling and
    inference that practitioners must consider when choosing between them.
    We provide relevant background contextualizing each approach and study
    their properties in a variety of scenarios, making recommendations for
    use based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

Spatial data are data that have some sort of spatial index --usually
this index is specified via coordinates. De Gruijter and Ter Braak
(1990) and Brus and DeGruijter (1993) give early comparisons of
design-based and model-based approaches for spatial data, quashing the
belief that design-based approaches could not be used for spatially
correlated data. Since then, there have been several general comparisons
between design-based and model-based approaches for spatial data (Brus
and De Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008). Cooper (2006)
reviews the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. In addition to Cooper (2006), there has been substantial
research and development into estimators that use both design-based and
model-based principles (see e.g., Sterba (2009) and Cicchitelli and
Montanari (2012), and for Bayesian approaches, see Chan-Golston et al.
(2020) and Hofman and Brus (2021)).

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less than
parameter estimates from samples lacking spatial balance (Barabesi and
Franceschi, 2011; Benedetti et al., 2017; Grafström and Lundström, 2013;
Robertson et al., 2013; Stevens and Olsen, 2004; Wang et al., 2013). The
first spatially balanced sampling algorithm to see widespread use was
the generalized random tessellation stratified (GRTS) algorithm (Stevens
and Olsen, 2004). After the GRTS algorithm was developed, several other
spatially balanced sampling algorithms emerged, including stratified
sampling with compact geographical strata Walvoort et al. (2010), the
local pivotal method (Grafström et al., 2012; Grafström and Matei,
2018), spatially correlated Poisson sampling (Grafström, 2012), balanced
acceptance sampling (Robertson et al., 2013), within-sample-distance
sampling (Benedetti and Piersimoni, 2017), and Halton iterative
partitioning sampling (Robertson et al., 2018). In this manuscript, we
select spatially balanced samples using the GRTS algorithm because it is
readily available in the \texttt{spsurvey} \textbf{\textsf{R}} package
(Dumelle et al., 2022) and naturally accommodates finite and infinite
sampling frames, unequal inclusion probabilities, and replacement units.
Replacement units are additional population units that can be sampled
when a population unit originally selected can no longer be sampled. A
couple reasons why an originally selected site can no longer be sampled
include its location being physically inaccessible or on private land
that the researcher does not have permission to access.

The GRTS algorithm selects samples by utilizing a particular mapping
between two-dimensional and one-dimensional space that preserves
proximity relationships. First the bounding box of the domain is split
up into four distinct, equally sized squares called level-one cells.
Each level-one cell is randomly assigned a level-one address of 0, 1, 2,
or 3. The set of level-one cells is denoted by \(\mathcal{A}_1\) and
defined as \(\mathcal{A}_1 \equiv \{a_1: a_1 = 0, 1, 2, 3\}\). Within
each level-one cell, the inclusion probability for each population unit
is summed, and if any of these sums exceed one, a second level of cells
is added. Then each level-one cell is split into four distinct, equally
sized squares called level-two cells. Each level-two cell is randomly
assigned a level-two address of 0, 1, 2, or 3. The set of level-two
cells is denoted by \(\mathcal{A}_2\) and defined as
\(\mathcal{A}_2 \equiv \{a_1a_2: a_1 = 0, 1, 2, 3; a_2 = 0, 1, 2, 3\}\).
The inclusion probabilities within each level-two cell are summed, and
if any of these sums exceed one, a third level of cells is added. This
process continues for \(k\) steps, until all level-\(k\) cells have
inclusion probability sums no larger than one. Then
\(\mathcal{A}_k \equiv \{a_1...a_k : a_1 = 0, 1, 2, 3; ...; a_k = 0, 1, 2, 3\}\).

After determining \(\mathcal{A}_k\), it is placed into hierarchical
order. Hierarchical order is a numeric order that first sorts
\(\mathcal{A}_k\) by the level-one addresses from smallest to largest,
then sorts \(\mathcal{A}_k\) by the level-two addresses from smallest to
largest, and so on. For example, \(\mathcal{A}_2\) in hierarchical order
is the set
\linebreak \(\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}\).
Because hierarchical ordering sorts by level-one cells, then level-two
cells, and so on, population units that have similar hierarchical
addresses tend to be nearby one another in space. Next each population
unit is mapped to a one-dimensional line in hierarchical order where
each population unit's inclusion probability equals its line-length. If
a level-\(k\) cell has multiple population units in it, they are
randomly placed within the cell's respective line segment. A uniform
random variable is then simulated in \([0, 1]\) and a systematic sample
is selected on the line, yielding \(n\) sample points for a sample size
\(n\). Each of these sample points falls on some population unit's line
segment, and thus that population unit is selected in the sample. For
further details regarding the GRTS algorithm, see Stevens and Olsen
(2004).

After selecting a sample and collecting data, unbiased estimates of
population means and totals can be obtained using the Horvitz-Thompson
estimator (Horvitz and Thompson, 1952). If \(\tau\) is a population
total, the Horvitz-Thompson estimator for \(\tau\), denoted by
\(\hat{\tau}_{ht}\), is is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n z_i \pi_i^{-1},
\end{align} where \(z_i\) is the value of the \(i\)th population unit in
the sample, \(\pi_i\) is the inclusion probability of the \(i\)th
population unit in the sample, and \(n\) is the sample size. An estimate
of the population mean is obtained by dividing \(\hat{\tau}_{ht}\) by
\(N\), the number of population units.

@@

Finite population block kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- for more details, see Ver Hoef (2008). Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

FPBK assumes \(\bm{\delta}\) in Equation\(~\)\ref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the separation vector (e.g., distance) between population units),
isotropic (independent of direction), and decay with distance between
population units (Cressie, 1993). Henceforth, it is implied that we have
made these same assumptions regarding \(\bm{\delta}\). Chiles and
Delfiner (1999), pp.~80-93 discuss covariance functions that are not
second-order stationary, not isotropic, or not either. A variety of
flexible covariance functions can be used to model \(\bm{\delta}\)
(Cressie, 1993); one example is the exponential covariance function
(Cressie (1993) provides a thorough list of spatial covariance
functions). The \(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter that
quantifies the spatially dependent variability, \(\sigma^2_{2}\) is the
variance parameter the quantifies that spatially independent
variability, \(\phi\) is the distance parameter that measures the
distance-decay rate of the covariance, and \(h_{i,j}\) is the Euclidean
distance between population units \(i\) and \(j\). In geostatistical
literature, \(\sigma^2_{1}\) is often called the partial sill,
\(\sigma^2_{2}\) is often called the nugget, and \(\phi\) is often
called the range.

The parameters in Equation\(~\)\ref{equation:Zmarginal} can be estimated
using a variety of techniques, but we focus on using restricted maximum
likelihood (Harville, 1977; Patterson and Thompson, 1971; Wolfinger et
al., 1994). REML is preferred over maximum likelihood (ML) because ML
estimates can be badly biased for small sample sizes, due to the fact
that ML makes no adjustment for the simultaneous estimation of
\(\bm{\beta}\) and \(\bm{\delta}\) (Patterson and Thompson, 1971). Minus
twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation} where
\(\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s\)
and \(\bm{\Sigma}_{ss}\) is the covariance matrix of the sampled sites.
Minimizing Equation\(~\)\ref{equation:reml} yields
\(\bm{\hat{\delta}}_{reml}\), the REML estimates of \(\bm{\delta}\).
Then \(\bm{\hat{\beta}}_{reml}\), the REML estimate of \(\bm{\beta}\),
is given by
\((\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s\),
where \(\bm{\hat{\Sigma}}_{ss}\) is \(\bm{\Sigma}_{ss}\) evaluated at
\(\bm{\hat{\delta}}_{reml}\).

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (there were no
replicates). The ZMMI and Hg ppb data are shown as spatial maps and as
histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest
near the coasts, lowest in the Central United States, are relatively
symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest
in the Northeastern United States, lowest elsewhere, are skewed, and
have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate
spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram
quantifies the the halved average squared differences (semivariance) of
responses whose separation (distance) falls within some distance class.
The spatial semivariance is closely related to the spatial covariance,
and spatial semivariograms are often used to gauge the strength of
spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately
strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance
increases steadily with distance (meaning that observations nearby one
another tend to be more similar than observations far apart from one
another).

@@

Mean bias was nearly zero for all four sampling-inference combinations
in all 36 scenarios, so we omit a more detailed summary of those results
here. Tables for mean bias in all 36 simulation scenarios are provided
in the supporting information.

We define the relative RMSE as a ratio with numerator given by the RMSE
for a sampling-inference combination and the denominator given by the
RMSE for SRS-DB. Relative RMSEs for the random location layout are
provided in Fig. \ref{fig:rmspe_eff}. When there is no spatial
covariance (Fig. \ref{fig:rmspe_eff}, ``DRE\%: 0\%''), the four
sampling-inference combinations have approximately equal RMSE. In these
scenarios, using GRTS sampling or model-based inference does not
generally increase efficiency compared to SRS-DB. When there is spatial
covariance (Fig. \ref{fig:rmspe_eff}, ``DRE\%: 50\%'' and ``DRE\%:
90\%''), GRTS-MB tends to have the lowest RMSE, followed by GRTS-DB,
SRS-MB, and finally SRS-DB, though the difference in relative RMSE among
GRTS-MB, GRTS-DB, and SRS-MB is small. As the strength of spatial
covariance increases, the gap in RMSE between SRS-DB and the other
sampling-inference combinations widens. Finally we note that when there
is spatial covariance, SRS-MB has a much lower RMSE than SRS-DB,
suggesting that the lack of efficiency from SRS is largely mitigated by
model-based inference. These RMSE conclusions are similar to those
observed in the grid location layout, so we omit a figure and discussion
regarding the grid location layout here. Tables for RMSE in all 36
simulation scenarios are provided in the supporting information.

@@

95\% interval coverage for each of the four sampling-inference
combinations in the random location layout is shown in Fig.
\ref{fig:figconf}. Within each simulation scenario, all
sampling-inference combinations tend to have fairly similar interval
coverage, though when \(n = 50\) or \(n = 100\), GRTS-DB coverage is
usually a few percentage points lower than the other combinations, which
suggests that the local neighborhood variance estimate may be slightly
too small for small \(n\). Coverage in the normal response scenarios was
usually near 95\%, while coverage in the skewed response scenarios
usually varied from 90\% to 95\% but increased with the sample size. At
a sample size of 200, all four sampling-inference combinations had
approximately 95\% interval coverage in both response scenarios for all
dependent error proportions. These interval coverage conclusions are
similar to those observed in the grid location layout, so we omit a
figure and discussion regarding the grid location layout here. Tables
for interval coverage in all 36 simulation scenarios are provided in the
supporting information.

@@

Mean bias was nearly zero for all four sampling-inference combinations
in all six scenarios, so we omit a more detailed summary of those
results here. Tables for mean bias in all six simulations scenarios are
provided in the supporting information.

The relative RMSE of both ZMMI (symmetric response) and Hg ppb (skewed
response) for all four sampling-inference combinations are shown in
Fig.\(~\)\ref{fig:data_rmspe_eff}. GRTS-MB has the lowest RMSE, followed
by GRTS-DB, SRS-MB, and then SRS-DB. The difference in RMSE among
GRTS-MB and GRTS-DB tends to be quite small. When \(n = 50\), SRS-MB
RMSE is approximately evenly between IRS-DB RMSE and GRTS-MB RMSE, but
for the larger sample sizes (\(n = 100\), \(n = 200\)), SRS-MB RMSE is
closer to GRTS-MB RMSE. Lastly we note that GRTS-MB, GRTS-DB, and SRS-MB
all have noticeably lower RMSE than SRS-DB. Tables for RMSE in all six
simulations scenarios are provided in the supporting information.

@@

We would like to thank the editors and anonymous reviewers for hard work
and time spent providing us with thoughtful, valuable feedback that
which greatly improved the manuscript.

>>>>

  \begin{abstract}
  \begin{enumerate}
  \def\labelenumi{\arabic{enumi}.}
  \tightlist
  \item
    The design-based and model-based approaches to frequentist statistical
    inference rest on fundamentally different foundations. In the
    design-based approach, inference relies on random sampling. In the
    model-based approach, inference relies on distributional assumptions.
    We compare the approaches for finite population spatial data.
  \item
    We provide relevant background for the design-based and model-based
    approaches and then study their performance using simulated and real
    data from the United States Environmental Protection Agency's 2012
    National Lakes Assessment. A variety of sample sizes, location
    layouts, dependence structures, and response types are considered. The
    population mean is the parameter of interest and performance is
    measured using statistics like bias, squared error, and interval
    coverage.
  \item
    When studying the simulated and real data, we found that regardless of
    the strength of spatial dependence in the data, the generalized random
    tessellation stratified (GRTS) algorithm, which explicitly
    incorporates spatial locations into sampling, tends to outperform the
    simple random sampling (SRS) algorithm, which does not explicitly
    incorporate spatial locations into sampling. We also found that
    model-based inference tends to outperform design-based inference, even
    for skewed data where the model-based distributional assumptions are
    violated. The performance gap between design-based inference and
    model-based inference is small when GRTS samples are used but large
    when SRS samples are used, suggesting that the sampling choice
    (whether to use GRTS or SRS) is most important when performing
    design-based inference.
  \item
    There are many benefits and drawbacks to the design-based and
    model-based approaches for finite population spatial sampling and
    inference that practitioners must consider when choosing between them.
    We provide relevant background contextualizing each approach and study
    their properties in a variety of scenarios, making recommendations for
    use based on the practitioner's goals.
  \end{enumerate}
  \end{abstract}

@@

Spatial data are data that have some sort of spatial index (usually
specified via coordinates). De Gruijter and Ter Braak (1990) and Brus
and DeGruijter (1993) give early comparisons of design-based and
model-based approaches for spatial data, quashing the belief that
design-based approaches could not be used for spatially correlated data.
Since then, there have been several general comparisons between
design-based and model-based approaches for spatial data (Brus and De
Gruijter, 1997; Brus, 2021; Ver Hoef, 2002, 2008). Cooper (2006) reviews
the two approaches in an ecological context before introducing a
``model-assisted'' variance estimator that combines aspects from each
approach. In addition to Cooper (2006), there has been substantial
research and development into estimators that use both design-based and
model-based principles (see e.g., Sterba (2009) and Cicchitelli and
Montanari (2012), and for Bayesian approaches, see Chan-Golston et al.
(2020) and Hofman and Brus (2021)).

@@

We previously mentioned that the design-based approach can be used to
select spatially balanced samples. Spatially balanced samples are useful
because parameter estimates from these samples tend to vary less than
parameter estimates from samples lacking spatial balance (Barabesi and
Franceschi, 2011; Benedetti et al., 2017; Grafström and Lundström, 2013;
Robertson et al., 2013; Stevens and Olsen, 2004; Wang et al., 2013). The
first spatially balanced sampling algorithm to see widespread use was
the generalized random tessellation stratified (GRTS) algorithm (Stevens
and Olsen, 2004). After the GRTS algorithm was developed, several other
spatially balanced sampling algorithms emerged, including stratified
sampling with compact geographical strata (Walvoort et al., 2010), the
local pivotal method (Grafström et al., 2012; Grafström and Matei,
2018), spatially correlated Poisson sampling (Grafström, 2012), balanced
acceptance sampling (Robertson et al., 2013), within-sample-distance
sampling (Benedetti and Piersimoni, 2017), and Halton iterative
partitioning sampling (Robertson et al., 2018). In this manuscript, we
select spatially balanced samples using the GRTS algorithm because it is
readily available in the \texttt{spsurvey} \textbf{\textsf{R}} package
(Dumelle et al., 2022) and naturally accommodates finite and infinite
sampling frames, unequal inclusion probabilities, and replacement units.
Replacement units are additional population units that can be sampled
when a population unit originally selected can no longer be sampled. A
couple of reasons why an originally selected site can no longer be
sampled include its location being physically inaccessible or it is on
private land that the researcher does not have permission to access.

The GRTS algorithm selects samples by utilizing a particular mapping
between two-dimensional and one-dimensional space that preserves
proximity relationships. First, the bounding box of the domain is split
up into four distinct, equally sized squares called level-one cells.
Each level-one cell is randomly assigned a level-one address of 0, 1, 2,
or 3. The set of level-one cells is denoted by \(\mathcal{A}_1\) and
defined as \(\mathcal{A}_1 \equiv \{a_1: a_1 = 0, 1, 2, 3\}\). Within
each level-one cell, the inclusion probability for each population unit
(which is pre-specified) is summed, and if any of these sums exceed one,
a second level of cells is added. Then each level-one cell is split into
four distinct, equally sized squares called level-two cells. Each
level-two cell is randomly assigned a level-two address of 0, 1, 2, or
3. The set of level-two cells is denoted by \(\mathcal{A}_2\) and
defined as
\(\mathcal{A}_2 \equiv \{a_1a_2: a_1 = 0, 1, 2, 3; a_2 = 0, 1, 2, 3\}\).
The inclusion probabilities within each level-two cell are summed, and
if any of these sums exceed one, a third level of cells is added. This
process continues for \(k\) steps, until all level-\(k\) cells have
inclusion probability sums no larger than one. Then
\(\mathcal{A}_k \equiv \{a_1...a_k : a_1 = 0, 1, 2, 3; ...; a_k = 0, 1, 2, 3\}\).
Figure\(~\)\ref{fig:grts} provides some intuition regarding the
assignment of level-one and level-two cells.

\begin{figure}
\centering
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/grts_level1.jpeg}
  \caption{Assignment of level-one cells to the spatial domain. Grey circles indicate population units.}
  \label{fig:grts_level1}
\end{subfigure}
\begin{subfigure}{0.49\textwidth}
  \centering
  \includegraphics[width = 1\linewidth]{figures/grts_level2.jpeg}
  \caption{Assignment of level-two cells to the spatial domain. Grey circles indicate population units.}
  \label{fig:grts_level2}
\end{subfigure}
\caption{Assignment of level-one and level-two cells to the spatial domain. In (a), each level-one cells is randomly given a level-one address of 0, 1, 2, or 3. In (b), each level-two cell within each level-one cell is randomly given a level-two address of 0, 1, 2, or 3.}
\label{fig:grts}
\end{figure}

After determining \(\mathcal{A}_k\), the set is placed into hierarchical
order. Hierarchical order is a numeric order that first sorts
\(\mathcal{A}_k\) by the level-one addresses from smallest to largest,
then sorts \(\mathcal{A}_k\) by the level-two addresses from smallest to
largest, and so on. For example, \(\mathcal{A}_2\) in hierarchical order
is the set
\linebreak \(\{00, 01, 02, 03, 10, ..., 13, 20, ..., 23, 30, ..., 33\}\).
Because hierarchical ordering sorts by level-one cells, then level-two
cells, and so on, population units that have similar hierarchical
addresses tend to be nearby one another in space. Next, each population
unit is mapped to a one-dimensional line in hierarchical order where
each population unit's inclusion probability equals its line-length. If
a level-\(k\) cell has multiple population units in it, they are
randomly placed within the cell's respective line segment. A uniform
random variable is then simulated in \([0, 1]\) and a systematic sample
is selected on the line, yielding \(n\) sample points for a sample size
\(n\). Each of these sample points falls on some population unit's line
segment, and thus that population unit is selected in the sample. For
further details regarding the GRTS algorithm, see Stevens and Olsen
(2004).

After selecting a sample and collecting data, unbiased estimates of
population means and totals can be obtained using the Horvitz-Thompson
estimator (Horvitz and Thompson, 1952). If \(\tau\) is a population
total, the Horvitz-Thompson estimator for \(\tau\), denoted by
\(\hat{\tau}_{ht}\), is given by \begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n z_i \pi_i^{-1},
\end{align} where \(z_i\) is the value of the \(i\)th population unit in
the sample, \(\pi_i\) is the inclusion probability of the \(i\)th
population unit in the sample, and \(n\) is the sample size. An estimate
of the population mean is obtained by dividing \(\hat{\tau}_{ht}\) by
\(N\), the number of population units.

@@

Finite population block kriging (FPBK) is a model-based approach that
expands the geostatistical Kriging framework to the finite population
setting (Ver Hoef, 2008). Instead of developing inference based on a
specific sampling design, we assume the data are generated by a spatial
stochastic process. We summarize some of the basic principles of FPBK
next -- see Ver Hoef (2008) for technical details and see Higham et al.
(2021) for an extension to cases of imperfect detection among population
units. Let
\({\mathbf{z} \equiv \{\text{z}(s_1), \text{z}(s_2), . . . , \text{z}(s_N) \}}\)
be an \(N \times 1\) response vector at locations \(s_1\), \(s_2\), . .
. , \(s_N\) that can be measured at the \(N\) population units. Suppose
we want to use a sample to predict some linear function of the response
variable, \(f(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}^\prime\) is a \(1 \times N\) vector of weights (e.g, the
population mean is represented by a weights vector whose elements all
equal \(1 / N\)). Denoting quantities that are part of the sampled
population units with a subscript \emph{s} and quantities that are part
of the unsampled population units with a subscript \emph{u}, let

@@

FPBK assumes \(\bm{\delta}\) in Equation\(~\)\eqref{equation:Zmarginal}
has mean-zero and a spatial dependence structure that can be modeled
using a covariance function. This covariance function is commonly
assumed to be non-negative, second-order stationary (depending only on
the separation vector (e.g., distance) between population units),
isotropic (independent of direction), and decays with distance between
population units (Cressie, 1993). Henceforth, it is implied that we have
made these same assumptions regarding \(\bm{\delta}\). Chiles and
Delfiner (1999), pp.~80-93 discuss covariance functions that are not
second-order stationary, not isotropic, or not either. A variety of
flexible covariance functions can be used to model \(\bm{\delta}\)
(Cressie, 1993); one example is the exponential covariance function
(Cressie (1993) provides a thorough list of spatial covariance
functions). The \(i,j\)th element of the exponential covariance matrix,
\(\mathop{\mathrm{{cov}}}(\bm{\delta})\), is \mbox{}
\begin{align}\label{equation:expcov}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = 
\begin{cases} 
\sigma^2_{1}\exp(-h_{i,j}/\phi) & h_{i,j} > 0 \\
\sigma^2_{1} + \sigma^2_2 & h_{i,j} = 0
\end{cases}
,
\end{align} where \(\sigma^2_{1}\) is the variance parameter that
quantifies the spatially dependent (correlated) variability,
\(\sigma^2_{2}\) is the variance parameter the quantifies that spatially
independent (not correlated) variability, \(\phi\) is the distance
parameter that measures the distance-decay rate of the covariance, and
\(h_{i,j}\) is the Euclidean distance between population units \(i\) and
\(j\). In geostatistical literature, \(\sigma^2_{1}\) is called the
partial sill, \(\sigma^2_{2}\) is called the nugget, and \(\phi\) is
called the range. We denote \(\bm{\theta}\) as the vector of covariance
parameters that composes \(\bm{\delta}\). In
Equation\(~\)\ref{equation:expcov},
\(\bm{\theta} = \{\sigma^2_1, \sigma^2_2, \phi \}\).

The parameters in Equation\(~\)\ref{equation:Zmarginal} can be estimated
using a variety of techniques, but we focus on using restricted maximum
likelihood (REML) (Harville, 1977; Patterson and Thompson, 1971;
Wolfinger et al., 1994). REML is preferred over maximum likelihood (ML)
because ML estimates can be badly biased for small sample sizes, due to
the fact that ML makes no adjustment for the simultaneous estimation of
\(\bm{\beta}\) and \(\bm{\delta}\) (Patterson and Thompson, 1971). Minus
twice the REML log-likelihood of the sampled sites is given by
\begin{equation}\label{equation:reml}
  \ln|\bm{\Sigma}| + (\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}})^T \bm{\Sigma}_{ss}^{-1}(\bm{z}_s - \bm{X}_s \bm{\tilde{\beta}}) + \ln|\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s| + (n - p) \ln(2 \pi) ,
\end{equation} where
\(\bm{\tilde{\beta}} = (\bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{X}_s)^{-1} \bm{X}_s^T \bm{\Sigma}_{ss}^{-1} \bm{z}_s\)
and \(\bm{\Sigma}_{ss}\) is the covariance matrix of the sampled sites.
Minimizing Equation\(~\)\ref{equation:reml} yields
\(\bm{\hat{\theta}}_{reml}\), the REML estimates of \(\bm{\delta}\).
Then \(\bm{\hat{\beta}}_{reml}\), the REML estimate of \(\bm{\beta}\),
is given by
\((\bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{X})^{-1} \bm{X}_s^T \bm{\hat{\Sigma}}_{ss}^{-1} \bm{z}_s\),
where \(\bm{\hat{\Sigma}}_{ss}\) is \(\bm{\Sigma}_{ss}\) evaluated at
\(\bm{\hat{\theta}}_{reml}\).

@@

The United States Environmental Protection Agency (USEPA), states, and
tribes periodically conduct National Aquatic Research Surveys (NARS) to
assess the water quality of various bodies of water in the contiguous
United States. One component of NARS is the National Lakes Assessment
(NLA), which measures various aspects of lake health and water quality.
We focus on analyzing zooplankton multi-metric indices (ZMMI) and
mercury concentrations in parts per billion (Hg ppb) from the 2012 NLA.
For ZMMI, data were collected at 1035 unique lakes. At less than 10\% of
lakes, two ZMMI replicates were collected. These were averaged for the
purposes of our study so that each lake had one measurement for ZMMI.
For Hg ppb, data were collected at 995 unique lakes (there were no
replicates). The ZMMI and Hg ppb data are shown as spatial maps and as
histograms in Figure \ref{fig:zmmi}. The ZMMI data tend to be highest
near the coasts, lowest in the Central United States, are relatively
symmetric, and have a mean of 55.05. The Hg ppb data tend to be highest
in the Northeastern United States, lowest elsewhere, are skewed, and
have a mean of 103.16 ppb. Also in Figure \ref{fig:zmmi} are separate
spatial semivariograms for ZMMI and Hg ppb. The spatial semivariogram
quantifies the halved average squared differences (semivariance) of
responses whose separation (distance) falls within some distance class.
The spatial semivariance is closely related to the spatial covariance,
and spatial semivariograms are often used to gauge the strength of
spatial dependence in data. Both ZMMI and Hg ppb seem to have moderately
strong spatial dependence (Figure \ref{fig:zmmi}), as the semivariance
increases steadily with distance (meaning that observations nearby one
another tend to be more similar than observations far apart from one
another).

@@

Mean bias is nearly zero for all four sampling-inference combinations in
all 36 scenarios, so we omit a more detailed summary of those results
here. Tables for mean bias in all 36 simulation scenarios are provided
in the supporting information.

We define the relative RMSE as a ratio with numerator given by the RMSE
for a sampling-inference combination and the denominator given by the
RMSE for SRS-DB. Relative RMSEs for the random location layout are
provided in Fig. \ref{fig:rmspe_eff}. When there is no spatial
covariance (Fig. \ref{fig:rmspe_eff}, ``DRE\%: 0\%''), the four
sampling-inference combinations have approximately equal RMSE. In these
scenarios, using GRTS sampling or model-based inference does not
generally increase efficiency compared to SRS-DB. When there is spatial
covariance (Fig. \ref{fig:rmspe_eff}, ``DRE\%: 50\%'' and ``DRE\%:
90\%''), GRTS-MB tends to have the lowest RMSE, followed by GRTS-DB,
SRS-MB, and finally SRS-DB. As the strength of spatial covariance
increases, the gap in RMSE between SRS-DB and the other
sampling-inference combinations widens. Finally we note that when there
is spatial covariance, SRS-MB has a much lower RMSE than SRS-DB,
suggesting that the lack of efficiency from SRS is largely mitigated by
model-based inference. These RMSE conclusions are similar to those
observed in the grid location layout, so we omit a figure and discussion
regarding the grid location layout here. Tables for RMSE in all 36
simulation scenarios are provided in the supporting information.

@@

95\% interval coverage for each of the four sampling-inference
combinations in the random location layout is shown in Fig.
\ref{fig:figconf}. Within each simulation scenario, all
sampling-inference combinations tend to have fairly similar interval
coverage, though when \(n = 50\) or \(n = 100\), GRTS-DB coverage is
usually a few percentage points lower than the other combinations, which
suggests that the local neighborhood variance estimate may be slightly
too small for small \(n\). Coverage in the normal response scenarios is
usually near 95\%, while coverage in the skewed response scenarios
usually varies from 90\% to 95\% but increases with the sample size. At
a sample size of 200, all four sampling-inference combinations have
approximately 95\% interval coverage in both response scenarios for all
dependent random error proportions. These interval coverage conclusions
are similar to those observed in the grid location layout, so we omit a
figure and discussion regarding the grid location layout here. Tables
for interval coverage in all 36 simulation scenarios are provided in the
supporting information.

@@

Mean bias is nearly zero for all four sampling-inference combinations in
all six scenarios, so we omit a more detailed summary of those results
here. Tables for mean bias in all six simulation scenarios are provided
in the supporting information.

The relative RMSE of both ZMMI (symmetric response) and Hg ppb (skewed
response) for all four sampling-inference combinations are shown in
Fig.\(~\)\ref{fig:data_rmspe_eff}. GRTS-MB has the lowest RMSE, followed
by GRTS-DB, SRS-MB, and then SRS-DB. The difference in RMSE among
GRTS-MB and GRTS-DB tends to be quite small. When \(n = 50\), SRS-MB
RMSE is approximately evenly between IRS-DB RMSE and GRTS-MB RMSE, but
for the larger sample sizes (\(n = 100\), \(n = 200\)), SRS-MB RMSE is
closer to GRTS-MB RMSE. Lastly we note that GRTS-MB, GRTS-DB, and SRS-MB
all have noticeably lower RMSE than SRS-DB. Tables for RMSE in all six
simulation scenarios are provided in the supporting information.

@@

We would like to thank the editors and anonymous reviewers for hard work
and time spent providing us with thoughtful, valuable feedback which
greatly improved the manuscript.
